Improving Dark Matter Axion Searches with Active Resonators

### 2015-05-12 paper/paper.tex
edited intro and rewrote experiment section

The signal power in microwave cavity experiments is proportional to a number of factors, including the local density of dark matter, the strength of the magnetic field, the volume of the cavity, and the resonant 
%loaded
quality factor $Q$ of the resonance being used:
%give equation
\begin{align}
P_{sig} = \eta g_{a\gamma\gamma}^2B_{ext}^2VC\frac{\rho_a}{m_a}\text{min}(Q_{loaded},Q_a)
\end{align}
The primary background is the thermal noise from the physical temperature of the cavity and the electronic noise from the first stage amplifier.  
The figure of merit for a microwave cavity experiment is the instantaneous axion signal to noise ratio (SNR).
%why emphasize instanteneous?
The SNR is given by the ratio of the signal power to the fluctuations in the noise power. We assume that we measure the signal for time $\tau$ with a frequency resolution equal to the width of the axion signal, $B_a$:
\begin{align}
SNR = \frac{P_{sig}}{\delta P_{noise}} = \frac{P_{sig}}{T_{noise}B_a}\sqrt{B_a \tau}
\end{align}

@@

Increasing the cavity $Q$ is one way to increase SNR in a microwave cavity experiment; the speed at which the cavity frequency can be tuned while remaining sensitive to a given axion photon coupling is linearly proportional to $Q$~\cite{Peng2000569}.
The presence of a strong magnetic field precludes the use of high-$Q$ superconducting cavities in these experiments, so the cavities are usually made of or coated with copper.
%add, ``...although R&D is being done with superconducting thing films on the walls...''
Loaded $Q$s of $10^5$ at 1 GHz have been achieved with copper cavities in axion experiments~\cite{Peng2000569}.  Higher frequency cavities typically have smaller $Q$s.
%mention anomalous skin depth limit, or scaling of Q with frequency as f^-2/3?
%The quality factor for a passive resonator depends solely on the frequency, geometry, and material of the resonator.
For a cylindrical copper cavity the Q scales as $Q \propto f^{-2/3}$....skin depth.

I present here a means of artificially increasing the $Q$ of the cavity resonance using active feedback in order to improve sensitivity to dark matter axions and increase the speed at which different axion masses can be tested.

@@

\begin{figure}[htbp]
\includegraphics[width=.5\textwidth]{figs/no_isolator}
\caption{Schematic of experimental setup}
\label{fig:schematic_of_prototype}
\end{figure}
In order to experimentally confirm the results of the analysis presented above, a microwave cavity with feedback loop was built and tested, as illustrated in Figure x. The enhanced cavity Q was varied by varying the attenuation in the loop and the output spectrum measured on a spectrum analyzer. A test tone was injected into a third port to measure the signal to noise ratio for various values of the enhanced Q. The phase was adjusted to provide maximum power output.\\

\begin{tabular}{|c|c|}
\hline
Parameters & Values \\ \hline
$f_0$ & 2.256$\times 10^9$ Hz \\
$Q_L$ & 1000 \\
$Q_0$ & ? \\
$\beta_1$ & ? \\
$\beta_2$ & ? \\
$T_{cav}$ & 300 K \\
$T_{amp}$ & 300 K \\
$G_a$ & ? dB \\
Averaging Time & (what is 25 avgs in secs?) \\
Time Delay & 2.4 $\mu$s \\
Bandwidth & ? wrote it down on a napkin? \\
\hline
\end{tabular}\\

As we injected a constant test tone and don't have a signal which scales with the enhanced Q, we expect only to observe an improvement in SNR of a factor of M.
The experiment was made with and without a delay line whose delay of $2.4\mu$seconds was a factor of 100 greater than the coherence time of the cavity. The improvement in SNR vs. enhanced Q is shown in Figure xx, for the delay line in and for no delay line. The improvement scales as expected for both cases.

>>>>

The signal power in microwave cavity experiments is proportional to a number of factors, including the local density of dark matter, the strength of the magnetic field, the volume of the cavity, and the resonant 
%loaded
quality factor $Q$ of the resonance being used \cite{TODOCITEEQN}:
%give equation
\begin{align}
P_{sig} = \eta g_{a\gamma\gamma}^2B_{ext}^2VC\frac{\rho_a}{m_a}\text{min}(Q_{loaded},Q_a)
\end{align}
where $\eta$ is the coupling between the cavity and output, $g_{a\gamma\gamma}$ is the axion-photon coupling, $B$ is the applied magnetic field, $V$ is the colume of the cavity, $C$ is the "form factor", or overlap between the cavity resonant mode and the axion field, $\rho_a$ is the local density of dark matter axions, $m_a$ is the axion mass, $Q_{loaded}$ is the loaded Q of the cavity resonant mode, $Q_a$ is the thermal width of the axion signal, taken to be $10^6$ or higher.
The primary background is the thermal noise from the physical temperature of the cavity and the electronic noise from the first stage amplifier.  
The figure of merit for a microwave cavity experiment is the instantaneous axion signal to noise ratio (SNR).
%why emphasize instanteneous?
The SNR is given by the ratio of the signal power to the fluctuations in the noise power. We assume that we measure the signal for time $\tau$ with a frequency resolution equal to the width of the axion signal, $B_a$:
\begin{align}
SNR = \frac{P_{sig}}{\delta P_{noise}} = \frac{P_{sig}}{T_{noise}}\sqrt{B_a \tau}
\end{align}

@@

Increasing the cavity $Q$ is one way to increase SNR in a microwave cavity experiment; the speed at which the cavity frequency can be tuned while remaining sensitive to a given axion photon coupling is linearly proportional to $Q$~\cite{Peng2000569}.
The presence of a strong magnetic field precludes the use of high-$Q$ superconducting cavities in these experiments, so the cavities are usually made of or coated with copper.
%add, ``...although R&D is being done with superconducting thing films on the walls...''
Loaded $Q$s of $10^5$ at 1 GHz have been achieved with copper cavities in axion experiments~\cite{Peng2000569}.  The Q of a cavity scales as $Q \propto f^{-2/3}$ due to skin depth effects \cite{TODOCITETHIS}, so that higher frequency cavities typically have smaller $Q$s.  This has the effect of making higher mass axions more challenging to search for.
%mention anomalous skin depth limit, or scaling of Q with frequency as f^-2/3?
%The quality factor for a passive resonator depends solely on the frequency, geometry, and material of the resonator.
%For a cylindrical copper cavity the Q scales as $Q \propto f^{-2/3}$....skin depth.

We present here a means of artificially increasing the $Q$ of the cavity resonance using active feedback in order to improve sensitivity to dark matter axions and increase the speed at which different axion masses can be tested.

@@

\begin{figure}[htbp]
\includegraphics[width=.5\textwidth]{figs/no_isolator}
\caption{Schematic of experimental setup}
\label{fig:schematic_of_prototype}
\end{figure}

Here we experimentally confirm the results of the analysis presented above.
A microwave cavity with a feedback loop was built as in Fig.~\ref{FIG:??}.
The axion signal was simulated with a standard signal generator fed into a weakly coupled port in the cavity.
The active Q of the system could be changed by modifying the phase and attenuation of the system.
A network analyzer attached ******* was used to measure the active Q of the system.
A specturm analyzer measured a fraction of the power in the feedback loop to measure signal and noise power. 
The relevant parameters of the setup are summarized in Table~\ref{TAB:PARAMS}.

The signal to noise enhancement was measured as a function of active Q for three different delays in the feedback loop.
First, the loop gain of the system was set by adjustion the variable attenuator.  
***ANA MAKE SURE THIS IS THE RIGHT PROCEDURE***
Then the phase shifter was adjusted to maximize the Q measured in the network analyzer.
This Q was recorded as the active Q.
With the signal generator off, the power at the center of the Q was measured with the spectrum analyzer.
This was recorded as the noise power.
The signal generator was then turned on, the frequency tuned to correspond to the center of the resonance, and the power was measured and recorded as the signal power. 
The variable attenuation was then decreased slightly and the procedure was repeated for the new, higher, active Q until the loop gain was high enough that the system became unstable and began to oscillate.

\begin{tabular}{|c|c|}
\hline
Parameters & Values \\ \hline
$f_0$ & 2.256$\times 10^9$ Hz \\
$Q_L$ & 1000 \\
$Q_0$ & ? \\
$\beta_1$ & ? \\
$\beta_2$ & ? \\
$T_{cav}$ & 300 K \\
$T_{amp}$ & 300 K \\
$G_a$ & ? dB \\
Averaging Time & (what is 25 avgs in secs?) \\
Time Delay & 2.4 $\mu$s \\
Bandwidth & ? wrote it down on a napkin? \\
\hline
\label{TAB:PARAMS}
\end{tabular}\\

The procedure was followed for three time delays in the feedback loop.  
The noise power and signal power are shown in Fig.~\ref{???}.
The ratio of the two, SNR, is shown in Fig.~\ref{??}.
As expected, both the signal and noise scale quadratically for the shortest delay, while the noise scales only linearly for the delay greater than $\frac{Q}{f}$.
For the intermediate delay, the transition in noise scaling from linear to quadratic is seen in the region where $t_{\mathrm{delay}}\simeq\frac{Q}{f}$. 
Thus the SNR is seen to increase linearly only for delays greater than $\frac{Q_{\mathrm{active}}}{f}$.
In the longest delay, a Q enchancement of **** could be achieved, with a proportional gain in SNR.
